\subsection{\elem{GROUP}}
\elem{GROUP} contains the time series data itself. A service providing time series \textbf{MUST} provide this element.

The \elem{GROUP} is identified with:
\begin{description}
\item ucd=''timeseries'' \todo{Please give input here}
\item utype =''timeseries'' \todo{Please give input here}
\end{description}

The \elem{GROUP} contains:
\begin{description}
     \item[\elem{name}] The name of the group. This attribute is mandatory and it should be set to \verb|timeseries|. 
     \item[\elem{DESCRIPTION}] Human readable text describing the time series. This element is mandatory. 
     \item[\elem{FIELDref}] One or several \elem{FIELDref} (or \elem{PARAMref}) to reference to the \elem{TIMESYS} element. At least one such reference MUST be provided.
     \item[\elem{FIELDref}] One or several \elem{FIELDref} (or \elem{PARAMref}) to reference to the \elem{COOSYS}, \elem{FILTERSYS} as applicable. 
     \item One or several PARAM or PARAMref. This will describe the BAND for instance. While PARAM is usually defining valid for a whole table in this case it will be valid only for the GROUP.  
     \item[\elem{SELECT}] One SELECT attribute. This element provides a way to reference rows instead of columns. This attribute is particular important for cases in which the magnitudes given in a column are not corresponding to the same band and the name of the band is given in a different column. \todo{I hope this is clear, case of ZTF for instance.}
\end{description}

Let us consider the simplest time series case possible: a table containing only three columns \emph{time, magnitude and magnitude error}, the minimum GROUP would be:

\begingroup
\input{vot-ex1.tex}
\endgroup

In this example we have assumened some metadata for \elem{TIMESYS} and \elem{FILTERSYS} and have included to show a full example case. While for the time and the magnitude reference implies an earlier definition of the elements \elem{TIMESYS} and \elem{FILTERSYS} this is not the case for the error. The error in the magnitude references the name of the column which will be repeated in the \elem{FIELD} element directly of the \elem{TABLE}.

We could argue that the \elem{GROUP} is not needed in this case, but the idea is to have a common way to representing time series, including the more complicated cases. 

\newpage 
Let us consider now consider a more complicated case, in which the time series also includes a column to state the name of the filter for wich the magnitudes are given, i.~e. a table containing four columns \emph{time, magnitude and magnitude error, filter name}. In those cases this complicates the annotation and the need of being able to reference rows (as well as columns) becomes clear.

\begingroup
\input{vot-ex2.tex}
\endgroup

To avoid referencing columns, which would make things a bit more complicated, we propose to create different \elem{TABLE} elements, one for each filter. This, although reduntant in the definition of some of the columns, will ease combining timeseries from the client side. 

%Alternative 2: Use the GROUP name=key thing and a different table to define the relation between elements. This 
%Alternative 3: Another possible solution would be to create a new XMLDATA serialization which would allow to tag rows, as proposed in Appendix A.8 of \cite{VOTable1.4}. But wouldn't that break the interoperability? 

